Global Life Expectancy: Unraveling Health and Economic Determinants (2000-2015)

### 2024-12-01 paper/paper.qmd
initial

The primary goal of this paper is to determine which factors play a statistically significant role in driving lower life expectancy values and to offer actionable insights based on country status. Using Multiple Linear Regression model, this study focuses on understanding the relationship between country status(developed or developing), continent, personal average expenditure on health based on GDP, percentage of government expenditure on health, income composition of resources and year of schooling in predicting life expectancy across different countries. By identifying and analyzing these predictors, the study aims to highlight actionable aspects for policymakers to target in their efforts to improve population longevity effectively.

@@

The measurement process refers to how real-world factors—such as the status of a country, average personal health care investment, average year of education, geographic location of a country and social determinants (country's healthcare investment, human development index of income) — are translated into numerical entries representing life expectancy in a dataset. Each entry captures the average life expectancy of individuals in a specific country during a given year.

@@

**Continent(`Continent`)**: This is a categorical variable that classifies the geographic location of the countries. It is detected by the name of the country using rpackage `countrycode`. The data is stored as the name of the continent(‘Africa’, ‘Oceania’, ‘Asia’, ‘Europe’, ‘North America’, ‘South America’)

@@

The raw life expectancy data underwent a several cleaning steps to ensure it was accurate, consistent and ready for analysis. We first select and rename key variables from raw data to focus on relevant information. The key variables of interest in our analysis include Country, Year, Status of the country, Life expectancy, Percentage expenditure, Total expenditure, Income composition of resources and the year of education. Additionally, we develop a new predictor named "Continent" to monitor each nation's geographic data. To make the subsequent analysis easier, we then convert variables to the proper data types and eliminate the rows that have missing data values. To keep things neat, we organize the decimal for every piece of numerical data. The cleaned dataset was then saved as a Parquet file for efficient storage and further analysis.

@@

2.  **Continent(`Continent`)**: This is the categorical variable showing the geographic information of each country. Developed continents, for example Europe and North America, typically have well-established healthcare infrastructure, which are considered to contribute to higher life expectancy. Factors like clean water access, education resources, nutrition and lifestyle factors can significantly impact life expectancy and vary greatly across continents. @fig-continent reveals that Africa has the highest number of countries represented with the largest portion of the data. Other continents, America and Europe, have fewer countries included, with only 9 countries from Oceania.

3.  **Percentage Expenditure on Health (`PercentageExpenditure`)**: This variable reflects the expenditure allocated to health-related expenditures as a percentage of GDP per capita.

4.  **Total expenditure (`TotalExpenditure`)**: This variable shows the general government expenditure on health as a percentage of total government expenditure. This is considered as a reflection of economic situation of a country. A higher percentage indicates that a country places significant emphasis on healthcare as a key area of public investment.

5. **Income composition of resource (`IncomeComposition`)**: his is a numerical variable representing the income component of the Human Development Index (HDI).

6. **Schooling (`Schooling`)**: This is a numerical variable showing the number of education per person.

@@

# Count the number of unique countries for each Status
country_counts <- cleaned_expectancy %>%
  distinct(Country, Status) %>%  # Ensure each country is only counted once
  count(Status)  # Count the number of countries in each Status category

@@

# Plot2: Create a bar plot to show the distribution of countries by Status
numbercountry <- ggplot(country_counts, aes(x = Status, y = n, fill = Status)) +
  geom_bar(stat = "identity") +  # Use 'identity' to directly use the count values
  geom_text(aes(label = n), vjust = -0.5,size = 2.5) +  # Add count labels on top of bars
  labs(
    title = "Distribution of Country by Status",
    x = "Status (Developed vs Developing)",
    y = "Number of Countries"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Developed" = "skyblue", "Developing" = "darkblue")) +
  theme(
    legend.position = "none", 
  plot.title = element_text(size = 7,face = "bold"))

@@

```{r}
#| label: fig-percentage
#| fig-cap: "Distribution of expenditure allocated to health-related expenditures as a percentage of GDP per capita. Africa has the highest number of countries represented, followed by Asia among all 5 continents. The distribution of count of data aslo shows that a significant portion of the data is focused on African and Asian nations.  "
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{r}
#| label: fig-income
#| fig-cap: "Distribution of contient. Africa has the highest number of countries represented, followed by Asia among all 5 continents. The distribution of count of data aslo shows that a significant portion of the data is focused on African and Asian nations.  "
#| echo: false
#| eval: true
#| warning: false
#| message: false

```{r}
#| label: fig-schooling
#| fig-cap: "Distribution of contient. Africa has the highest number of countries represented, followed by Asia among all 5 continents. The distribution of count of data aslo shows that a significant portion of the data is focused on African and Asian nations.  "
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```

$$overall\_time\_seconds_i​=\beta_0​+s(age_i​)+\beta_1⋅gender_i​+s(race\_count_i​​)+ \beta_2⋅iaaf\_category​​ + ϵ_i​$$

@@

-   $overall\_time\_seconds_i$ is the marathon finishing time in seconds for the i-th runner.

-   $s(age_i)$ is a smooth function of the runner’s age at the time of the race.

-   $gender_i$ is the runner's gender (treated as a categorical variable).

-   $s(race\_count_i)$ is a smooth function of the total number of NYRR races the runner has participated in.

-   $iaaf\_category_i$​ is the country the runner is from under IAAF standards.

@@

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```
We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

@@

@@

>>>>

The primary goal of this paper is to determine which factors play a statistically significant role in driving lower life expectancy values and to offer actionable insights based on country status. Using Multiple Linear Regression model, this study focuses on understanding the relationship between country status(developed or developing), personal average expenditure on health based on GDP, percentage of government expenditure on health, income composition of resources and year of schooling in predicting life expectancy across different countries. By identifying and analyzing these predictors, the study aims to highlight actionable aspects for policymakers to target in their efforts to improve population longevity effectively.

@@

The measurement process refers to how real-world factors—such as the status of a country, average personal health care investment, average year of education and social determinants (country's healthcare investment, human development index of income) — are translated into numerical entries representing life expectancy in a dataset. Each entry captures the average life expectancy of individuals in a specific country during a given year.

@@

@@

The raw life expectancy data underwent a several cleaning steps to ensure it was accurate, consistent and ready for analysis. We first select and rename key variables from raw data to focus on relevant information. The key variables of interest in our analysis include Country, Year, Status of the country, Life expectancy, Percentage expenditure, Total expenditure, Income composition of resources and the year of education. To make the subsequent analysis easier, we then convert variables to the proper data types and eliminate the rows that have missing data values. To keep things neat, we organize the decimal for every piece of numerical data. The cleaned dataset was then saved as a Parquet file for efficient storage and further analysis.

@@

2.  **Percentage Expenditure on Health (`PercentageExpenditure`)**: This variable reflects the expenditure allocated to health-related expenditures as a percentage of GDP per capita. This percentage is generally expected to fall between 0% and 100%, where 100% would indicate that the total GDP per capita is allocated to healthcare. @fig-percentage shows that in most of the years from 2000 to 2015, most of the countries have the individual health expenditure higher than their GDP. It is noticeable that some developing countries have individual health expenditure less than 20% of the GDP.

3.  **Total expenditure (`TotalExpenditure`)**: This variable shows the general government expenditure on health as a percentage of total government expenditure. This is considered as a reflection of economic situation of a country. A higher percentage indicates that a country places significant emphasis on healthcare as a key area of public investment.

4.  **Income composition of resource (`IncomeComposition`)**: his is a numerical variable representing the income component of the Human Development Index (HDI).

5.  **Schooling (`Schooling`)**: This is a numerical variable showing the year of receiving education. @fig-schooling is a combination of box plot and violin plot showing the year of schooling differs by the status. In development countries, the average number of years of schooling is approximately 16 years, with a narrow distribution suggesting that the educational attainment in these countries is more consistent. In contrast, in developing countries, the average number of years od schooling is around 11 years. With a much wider spread and significant more outlier, the education situation in developing countries is less stable.

@@

# Count the number of unique countries for each Status
country_counts <- cleaned_expectancy %>%
  # Ensure each country is only counted once
  distinct(Country, Status) %>%  
   # Count the number of countries in each Status category
  count(Status)

@@

# Plot2: Create a bar plot to show the distribution of countries by Status
numbercountry <- ggplot(country_counts, aes(x = Status, y = n, fill = Status)) +
   # Use 'identity' to directly use the count values
  geom_bar(stat = "identity") + 
   # Add count labels on top of bars
  geom_text(aes(label = n), vjust = -0.5,size = 2.5) +  
  labs(
    title = "Distribution of Country by Status",
    x = "Status (Developed vs Developing)",
    y = "Number of Countries"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("Developed" = "skyblue", "Developing" = "darkblue")) +
  theme(
    legend.position = "none", 
  plot.title = element_text(size = 7,face = "bold"))

@@

```{r}
#| label: fig-percentage
#| fig-cap: "Distribution of Health Expenditure ranges by country status. With the orange color showing the developing and blue showing developed, most of the country has more than 100% of their GDP per capita allocated to health-related expenditures."
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Define percentage ranges
ranges <- c("0%-20%", "20%-30%", "30%-50%", "50%-100%", ">100%")

# Create a new variable to classify the ranges
expenditure_ranges <- cleaned_expectancy %>%
  mutate(
    ExpenditureRange = case_when(
      PercentageExpenditure <= 20 ~ "0%-20%",
      PercentageExpenditure > 20 & PercentageExpenditure <= 30 ~ "20%-30%",
      PercentageExpenditure > 30 & PercentageExpenditure <= 50 ~ "30%-50%",
      PercentageExpenditure > 50 & PercentageExpenditure <= 100 ~ "50%-100%",
      PercentageExpenditure > 100 ~ ">100%"
    )
  ) %>%
  # Ensure the right order
  mutate(ExpenditureRange = factor(ExpenditureRange, levels = ranges))

# Plot the histogram of expenditure ranges separated by status
ggplot(expenditure_ranges, aes(x = ExpenditureRange, fill = Status)) +
  geom_bar(stat = "count", position = "dodge", color = "white", alpha = 0.7) +
  # Adds count on top of each bin
  geom_text(stat='count', aes(label = ..count..), position = position_dodge(width = 0.8), vjust = - 0.5, size = 2.5) +
  scale_fill_manual(values = c("Developed" = "blue", "Developing" = "orange")) +
  labs(
    title = "Distribution of Health Expenditure Ranges by Country Status",
    x = "Health Expenditure Range (% of GDP per Capita)",
    y = "Count",
    fill = "Status"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1), # Adjusts the x-axis labels for clarity
    plot.title = element_text(size = 10,face = "bold")) 
```

# Define percentage ranges
ranges <- c("0%-2%", "2%-4%", "4%-6%", "6%-8%", "8%-10%", "10%-12%","12%-14%", ">14%")

# Create a new variable to classify the ranges
countryexpenditure <- cleaned_expectancy %>%
  mutate(
    TotalRange = case_when(
      TotalExpenditure <= 2 ~ "0%-2%",
      TotalExpenditure > 2 & TotalExpenditure <= 4 ~ "2%-4%",
      TotalExpenditure > 4 & TotalExpenditure <= 6 ~ "4%-6%",
      TotalExpenditure > 6 & TotalExpenditure <= 8 ~ "6%-8%",
      TotalExpenditure > 8 & TotalExpenditure <= 10 ~ "8%-10%",
      TotalExpenditure > 10 & TotalExpenditure <= 12 ~ "10%-12%",
      TotalExpenditure > 12 & TotalExpenditure <= 14 ~ "12%-14%",
      TotalExpenditure > 14 ~ ">14%"
    )
  ) %>%
  # Ensure the right order
  mutate(TotalRange = factor(TotalRange, levels = ranges))

summary(countryexpenditure$TotalRange)
table(countryexpenditure$TotalRange)

# Plot the histogram of expenditure ranges separated by status
ggplot(countryexpenditure, aes(x = TotalRange, fill = Status)) +
  geom_bar(stat = "count", position = "dodge", color = "white", alpha = 0.7) +
  # Adds count on top of each bin
  geom_text(stat='count', aes(label = ..count..), position = position_dodge(width = 0.8), vjust = - 0.5, size = 2.5) +
  scale_fill_manual(values = c("Developed" = "lightblue", "Developing" = "salmon")) +
  labs(
    title = "Distribution of Government Expenditure on Health as Percentage Range by Country Status",
    x = "Government Expenditure Range (%)",
    y = "Count",
    fill = "Status"
  ) +
  theme_minimal() +
  theme(
    # Sidebar for the legend
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1), # Adjusts the x-axis labels for clarity
    plot.title = element_text(size = 10,face = "bold"))

@@

```{r}
#| label: fig-income
#| fig-cap: "Combined Box plot and Violin plot of income component of the Human Development Index (HDI) showing the variance and the average of the income component under status of different statuses."
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Create a combined boxplot and violin plot
ggplot(cleaned_expectancy, aes(x = Status, y = IncomeComposition, fill = Status)) +
  # Violin plot shows distribution
  geom_violin(trim = FALSE, alpha = 0.5) + 
  # Boxplot shows summary statistics
  geom_boxplot(width = 0.2, alpha = 0.7, outlier.color = "red", outlier.shape = 16) + 
  # Highlight mean
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "darkblue") + 
  labs(
    title = "Income Composition of Resources by Country Status",
    x = "Country Status",
    y = "Income Composition (HDI Contribution)"
  ) +
  scale_fill_manual(values = c("Developed" = "skyblue", "Developing" = "salmon")) +
  theme_minimal() +
  theme(
    # Remove legend as x-axis indicates status
    legend.position = "none", 
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )
```

```{r}
#| label: fig-schooling
#| fig-cap: "Distribution of school separate by Status. In developed countries, the average number of years of schooling is approximately 16 years, typically following a bell-shaped distribution. In contrast, in developing countries, the average number of years of schooling is around 11 years. However, the variance in the years of schooling is significantly larger."
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Create a combined boxplot and violin plot
ggplot(cleaned_expectancy, aes(x = Status, y = Schooling, fill = Status)) +
  # Violin plot shows distribution
  geom_violin(trim = FALSE, alpha = 0.5) + 
  # Boxplot shows summary statistics
  geom_boxplot(width = 0.2, alpha = 0.7, outlier.color = "red", outlier.shape = 16) + 
  # Highlight mean
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "darkblue") + 
  labs(
    title = "Schooling by Country Status",
    x = "Country Status",
    y = "Year of Schooling"
  ) +
  scale_fill_manual(values = c("Developed" = "skyblue", "Developing" = "pink")) +
  theme_minimal() +
  theme(
    # Remove legend as x-axis indicates status
    legend.position = "none", 
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )
```

@@

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

@@

-   $Lifeexpectancy\_developed$ is the life expectancy of developed country.

-   $Lifeexpectancy\_developing$ is the life expectancy of developing country.

-   $PercentageExpenditure$ is a smooth function of the runner’s age at the time of the race.

-   $Continent$ is

-   $TotalExpenditure$ is the runner's gender (treated as a categorical variable).

-   $IncomeComposition$ is a smooth function of the total number of NYRR races the runner has participated in.

-   $Schooling$​ is the country the runner is from under IAAF standards.

@@

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

@@

## Idealized Survey

**Survey: Understanding Life Expectancy and Influencing Factor**

Thank you for participating in this survey. This survey aims to gather insights into the factors influencing life expectancy, including health expenditure, education, ethnicity, government spending on healthcare, and income levels. Your responses will help us understand individual perspectives and experiences towards national policy. Participation is voluntary, and your answers will remain anonymous.

**Contact Information:** If you have any questions about the survey or the data collection process, please contact

**Survey Coordinator**: Yanfei Huang\
**Email**: yanfei.huang\@mail.utoronto.ca

**Section 1: Personal Health Expenditure**

**1.What percentage of your monthly income do you spend on healthcare (e.g., insurance, medications, doctor visits)?**

-   Less than 5%
-   5%–10%
-   10%–20%
-   More than 20%

**2.Do you or your household have health insurance coverage?**

-   Yes
-   No

**3. How frequently do you visit a healthcare professional(eg.doctors, nurse) in a year?**

-   0-1 times
-   2-5 times
-   6-10 times
-   More than 10 times

**Section 2: Education Background**

**4.What is the highest level of education you have completed?**

-   No formal education
-   Primary school
-   Secondary school
-   College or university degree
-   Postgraduate degree

**5.How many years of formal schooling have you completed?**

Please write the number: \_\_\_\_\_\_\_

**Section 3: Ethnicity and Geographic Factors**

**6.Which of the following best describes your ethnicity?**

-   African
-   Asian
-   European
-   Hispanic or Latino
-   Middle Eastern
-   Indigenous
-   Other (please specify): \_\_\_\_\_\_\_

**7.What's the distance to the nearest healthcare facility from your residence?**

- Less than 1 km 
- 1-5 km 
- More than 5 km

**Section 4: Government Expenditure on Healthcare**

**8.Are health services in your country subsidized by the govenment?**

-   Yes, fully subsidized
-   Yes, partially subsidized
-   No

**9. What is the approximate cost of your most recent healthcare visit (in local currency)?**

Please write the amount \_\_\_\_\_\_\_

**10.Do public healthcare facilities in your area provide all the services you need?**

-   Yes
-   No
-   Not applicable

**Section 5: Income and Development Factor**

**11.What is your approximate monthly household income after taxes?(in local currency)**

-   Less than 1,000
-   1,000 - 2,999
-   3,000 - 4,999
-   5,000 - 7,999
-   8,000 - 10,999
-   More than 11,000
-   Prefer not to say

**12.How would you describe your current financial situation?**

-   Struggling to make ends meet
-   Just getting by
-   Comfortable, but not wealthy
-   Financially secure
-   Wealthy
-   Prefer not to say

**13.How often do you have difficulty affording basic healthcare (medical visits, medications, etc.)?**

-   Never
-   Rarely
-   Sometimes
-   Often
-   Always
-   Prefer not to say

**Final Section**

Thank you for completing this survey! Your responses will help capture information about economic conditions, access to healthcare and education, and basic living standards, which are critical for understanding life expectancy predictors and the impact of socioeconomic factors on health outcomes.

@@

## Additional graph for analysis
